%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------


\documentclass{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{upgreek}
\usepackage{pig}

\newcommand{\hb}{\!:\!}
\newcommand{\lol}{\multimap}
\newcommand{\nts}[1]{\marginpar{\tiny #1}}
\newcommand{\stk}[1]{\begin{array}{@{}c@{}}#1\end{array}}
\newcommand{\stkl}[1]{\begin{array}{@{}l@{}}#1\end{array}}

\newtheorem{lem}{Lemma}
\newtheorem{defn}[lem]{Definition}
\newtheorem{thm}[lem]{Theorem}
\newtheorem{cor}[lem]{Corollary}
\newtheorem{fail}[lem]{Failure}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the 
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish, 
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers, 
                                  % short abstracts)

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Quadratisch. Praktisch. Gut.}
%\subtitle{Subtitle Text, if any}

\authorinfo{Witheld}
           {Witheld}
           {witheld}

\maketitle

\begin{abstract}
\end{abstract}

\category{D.3.2}{Programming Languages}{Formal Definitions and Theory}

% general terms are not compulsory anymore, 
% you may leave them out
%\terms
%term1, term2

\keywords
dependent types, cubical sets

\section{Introduction}

We define a dependent type theory, where values (including types, of
course) can be joined by \emph{paths}, conceptualised as continuous
functions over the interval $[0,1]$. This type theory has the
following properties:
\begin{description}
\item[canonicity] closed expressions compute to canonical values;
\item[substitutivity] paths between values yield
  computational coercions between types which depend on those values;
\item[decidability] type checking is algorithmic;
\item[extensionality] the notion of path between
  functions is pointwise.
\end{description}

The definition is similar in spirit to the \emph{Observational
Type Theory} of Altenkirch, McBride and
Swierstra~\cite{DBLP:conf/plpv/AltenkirchMS07},
but its construction is greatly simplified by the adoption of
technology from the \emph{Cubical Type Theory} of Bezem, T. Coquand
and Huber~\cite{DBLP:conf/types/BezemCH13}.

By careful engineering, the definition we present has the property
of \textbf{proof erasure}, meaning that closed expressions can be
given a run time behaviour which does not rely in any way on the
data contained in paths. However, this property is
fragile in that it relies on the choice to connect types and
values only structurally.

Our type theory does \emph{not} exhibit the either of the following
mutually incompatible properties:
\begin{description}
\item[proof obliviousness] all paths between the same endpoints are
  judgmentally equal (an extreme form of the `K axiom', better known
  as `uniqueness of identity proofs');
\item[univalence] arbitrary computational equivalences between types induce paths.
\end{description}
Rather, we stand at the crossroads between the two: we can extend our
system either by internalizing the choice that paths can only be
structural, doubling down on proof erasure, or by destroying that
choice with additional paths containing nontrivial data thus
rejecting proof erasure for a run time computational system which
computes the representation changes which implement type equivalences.


\section{Bidirectional Type Theory, Still in Eden}

\newcommand{\gs}{\;\mid\;}
\newcommand{\neu}{\underline}
\newcommand{\TYPE}{\bigstar}
\newcommand{\EC}{\mathcal{E}}

The syntax of our type theory has sorts (which are types of types)
including a top sort $\TYPE$,
terms (whose types are checked) embedding eliminations (whose types are
synthesized) embedding variables in turn. Variables are assigned types by contexts.
\[\begin{array}{rrl}
\star & ::= & \TYPE \gs \ldots \\
s,t,P,R,S,T & ::= & \star \gs \neu e \gs \ldots \\
e,f,E,F & ::= & x \gs s\hb S \gs \ldots \\
\Gamma & ::= & \EC \gs \Gamma,x\hb S \gs \ldots
\end{array}\]

The typing rules are presented via three judgment forms: checking,
synthesis and subtyping. We omit contexts from typing rules, adopting
the convention that the context at a rule's conclusion is distributed
to all its premises. We thus need only write the \emph{extension} of
a context with an entry for a fresh variable, $x\hb S\vdash \cdots$,
or a context entry lookup for an existing variable $\dashv x\hb S$.
This discipline not only saves space: it also helps us to ensure that
the type system is stable under substitution by ensuring that we
treat variables only in ways whch are compatible with how we treat
terms. Before we add any `features' to our type theory, we must have
at least the following.

\[\begin{array}{c}
\textsc{syn-var}\;
\Rule{\dashv x\hb S}
     {x\in S}
\qquad\qquad
\textsc{syn-cut}\;
\Rule{\TYPE\ni S\quad S\ni s}
     {s\hb S\in S}
\\
\textsc{chk-syn}\;
\Rule{e\in S\quad S\le T}
     {T\ni \neu{e}}
\\
\textsc{sub-refl}\;
\Axiom{T\le T}
\qquad\qquad
\textsc{sub-top}\;
\Axiom{\star \le \TYPE}
\end{array}\]

So, for variables we synthesize exactly the type assigned by the
context; for cuts, we synthesize exactly the type proposed in the
annotation (provided it is checkably a type and indeed the type of the
term). The reason why a term with a type annotation is called a `cut':
it is in some sense activated for computation by being labelled with
its type, indicating its functionality.
If we can synthesize the type of something, we may surely check
it by requiring the type we get to be a subtype of the type we need,
where subtyping is at the very least reflexive.

Terms are blessed with a contraction scheme: cut elimination. A cut
embedded directly back into the term syntax can do no computation and
thus needs no annotation.
\[
\textsc{red-}\upsilon\quad
\neu{t\hb T}\leadsto t
\]

As soon as we have computation, we need to know its impact on
typechecking. We have rules which allow us to precompute types before
checking and postcompute types after synthesis.
\[
\textsc{chk-pre}\;
\Rule{T\leadsto R \quad R\ni t}
     {T\ni t}
\qquad
\textsc{syn-post}\;
\Rule{e\in S \quad S\leadsto R}
     {e\in R}
\]

Let us add dependent pairs ($\Sigma$-types), by way of an example feature.
\[\begin{array}{c}
\textsc{chk-sg}\;
\Rule{\star\ni S\quad x\hb S\vdash \star\ni T[x]}
     {\star\ni (x\hb S)\times T[x]}
\qquad
\textsc{syn-1}\;
\Rule{e\in (x\hb S)\times T[x]}
     {e.1\in S}
\\
\textsc{chk-pair}\;
\Rule{S\ni s\quad T[s\hb S]\ni t}
     {(x\hb S)\times T[x]\ni s,t}
\qquad
\textsc{syn-2}\;
\Rule{e\in (x\hb S)\times T[x]}
     {e.2\in T[e.1]}
\\
\textsc{sub-sg}\;
\Rule{S\le S' \quad x\hb S\vdash T[x]\le T'[x]}
     {(x\hb S)\times T[x]\le(x\hb S')\times T'[x]}
\end{array}\]
Every sort is closed under the formation of
$\Sigma$-types. $\Sigma$-types
admit pairs, where the first component, once checked, is used to
compute the type of the second. By putting introduction forms in
the terms and elimination forms in the eliminations (hence the
names), we ensure that introduction forms can only be eliminated
at a cut which exposes the type at which activity is present.
Computation, indeed, goes by $\beta$-rules.
\[\begin{array}{c}
\textsc{red-$\beta$-1}\quad 
  (s,t:(x\hb S)\times T[x]).1 \leadsto s:S \\
\textsc{red-$\beta$-2}\quad 
  (s,t:(x\hb S)\times T[x]).2 \leadsto t:T[s\hb S]
\end{array}
\]
So, a computation labelled as being at a compound type `decays' to
subcomputations labelled at the components (or instances of them).
Cuts bubble outwards through the elimination syntax, decaying as they go,
until they reach the point where the result is embedded in a term,
whereupon the cut can be eliminated.

For function types, we see a similar pattern.
\[\begin{array}{c}
\textsc{chk-pi}\;
\Rule{\star\ni S\quad x\hb S\vdash \star\ni T[x]}
     {\star\ni (x\hb S)\to T[x]}
\\
\textsc{chk-lam}\;
\Rule{x\hb S\vdash T[x]\ni t[x]}
     {(x\hb S)\to T[x]\ni \lambda x.\,t[x]}
\\
\textsc{syn-app}\;
\Rule{f\in (x\hb S)\times T[x]\quad S\ni s}
     {f\:s \in T[s\hb S]}
\\
\textsc{sub-pi}\;
\Rule{S'\le S \quad x\hb S'\vdash T[x]\le T'[x]}
     {(x\hb S)\to T[x]\le(x\hb S')\to T'[x]}
\\
\textsc{red-}\beta\textsc{-app}\quad
(\lambda x.\,t[x]:(x\hb S)\times T[x])\:s \leadsto t[s\hb S]:T[s\hb S]
\end{array}\]

Let us add Booleans as a base type in every sort, but with large elimination.
\newcommand{\Two}{2}
\newcommand{\true}{\mathsf{t\!t}}
\newcommand{\false}{\mathsf{f\!f}}
\newcommand{\iffy}[4]{(\textsc{if}\:(#1.\,#2)\:#3\:#4)}
\[\begin{array}{c}
\textsc{chk-two}\;
\Axiom{\star\ni \Two}\qquad
\textsc{chk-tt}\;
\Axiom{\Two\ni \true}\qquad
\textsc{chk-ff}\;
\Axiom{\Two\ni \false}\\
\textsc{syn-if}\;
\Rule{e\in\Two\quad b\hb\Two\vdash \TYPE\ni P[b]\quad
      P[\true\hb\Two]\ni t\quad P[\false\hb\Two]\ni f}
     {e\:\iffy{b}{P[b]}{t}{f}\in P[e]}\\
\textsc{red-}\beta\textsc{-if-tt}\quad
(\true\hb\Two)\:\iffy{b}{P[b]}{t}{f} \leadsto t\hb P[\true\hb\Two] \\
\textsc{red-}\beta\textsc{-if-ff}\quad
(\false\hb\Two)\:\iffy{b}{P[b]}{t}{f} \leadsto f\hb P[\false\hb\Two] \\
\end{array}\]

We can also have empty and unit types.
\newcommand{\Zero}{\mathsf{0}}
\newcommand{\One}{\mathsf{1}}
\newcommand{\void}{\langle\rangle}
\newcommand{\magic}[1]{(\textsc{magic}\:#1)}
\[\begin{array}{c}
\textsc{chk-zero}\;
\Axiom{\star\ni \Zero}\qquad
\textsc{chk-one}\;
\Axiom{\star\ni \One}\\
\textsc{syn-magic}\;
\Rule{e\in\Zero\quad \TYPE\ni P}
     {e\:\magic{P}\in P}\qquad
\textsc{chk-void}\;
\Axiom{\One\ni \void}\\
\end{array}\]
We could add a dependent eliminator for $\One$, asserting that every
element is $\void$, or we could just add an $\eta$-rule.

\newcommand{\Type}[1]{\star^{#1}}

\textbf{datatypes of some sort?}

\newcommand{\jq}{\cong}
\textbf{introduce the judgmental equality} $\jq$


Equality is a congruence respecting $\leadsto$, but it is not
necessarily the least such. We may also consider type-directed
$\eta$-expansion rules, such as
\[\begin{array}{@{}l@{\;\;R=}l@{\;\Rightarrow\; R\;\ni\; r\jq\; }l@{}}
\textsc{one-$\eta$}&\One&\void \\
\textsc{sg-$\eta$}&(x\hb S)\times T[x])& (r\hb R).1, (r\hb R).2 \\
\textsc{pi-$\eta$}&(x\hb S)\to T[x])& \lambda x.\,(r\hb R)\:x\\
\end{array}\]


\section{Paths Abstracting Types Over the Interval}

The term `equality' is laden with philosophical baggage. Let us slough
off that burden and consider the idea of connecting things with paths.
There might very well be information in a path connecting things that
is not already in the things themselves: for example, a road route
connecting Geneva and Montreux will have to go around Lake Geneva one
way or the other. Of course, we shall need to know how to construct
paths and how we expect to be able to use them.

\newcommand{\pl}{0}
\newcommand{\pr}{1}
The key mechanism of cubical type theories is the idea of constructing
paths in a pointwise manner: we allow a parameter $i$ to vary
`continuously' between $\pl$ and $\pr$, and we give a construction
$t[i]$ valid at all $i$, thus exposing a connection between $t[\pl]$
and $t[\pr]$. Of course, paths are constructions, too, so we can form
paths-between-paths by abstracting over one parameter a construction
which then abstracts over a second: some $t[i,j]$ varying over a
square. Paths-between-paths-between-paths abstract over cubes, and so
on up the dimensions, hence `cubical' type theory.

Let us start with paths between types.
\newcommand{\tp}[1]{\frac{#1}{\;\;\;}}
\[\begin{array}{c}
\textsc{chk-typ}\;
\Rule{\star\le\star'\quad \star \ni S\quad\star\ni T}
     {\star' \ni S\tp{\star}T}
\\
\textsc{chk-typ-lam}\;
\Rule{\bar i\vdash \star \ni P[i]\quad \star\ni S\jq P[\pl]\quad
    \star\ni P[\pr]\jq T}
     {S\tp{\star}T \ni \lambda i.\,P[i]}
\end{array}\]
Note that the interval between $\pl$ and $\pr$ is not a type, so the
binding in context $\bar i$ is of a different form. We shall need a
separate judgment for \emph{points}, offering at least the following.
\newcommand{\pt}[1]{\textsc{pt}\;#1}
\[
\textsc{pt-$\pl$}\;\Axiom{\pt\pl}\qquad 
\textsc{pt-$\pr$}\;\Axiom{\pt\pr}\qquad
\textsc{pt-var}\;\Rule{\dashv \bar i}
                      {\pt i}
\]

Now we can explain how to project from a path at a given
point.
\[\begin{array}{c}
\textsc{syn-typ-app}\;
\Rule{E\in S\tp{\star}T\quad \pt a}
     {E\:a\in\star}
\\
\textsc{red-typ-$\beta$}\;
(\lambda i.\,P[i]:S\tp{\star}T)\:a\leadsto P[a]:\star
\end{array}\]

The above $\beta$-rule is enough for the evaluation of closed
expressions, but the equational theory of open terms can readily
be enriched by further computation rules, allowing projection
of \emph{endpoints} which are known, in that types of eliminations
can be reconstructed.
\[\begin{array}{c}
\textsc{red-typ-$\pl$}\;
  E\in S\tp\star T \;\Rightarrow\; E\:\pl\leadsto S\hb\star\\
\textsc{red-typ-$\pr$}\;
  E\in S\tp\star T \;\Rightarrow\; E\:\pr\leadsto T\hb\star\\
\end{array}\]
The critical pairs with $\beta$ induced by these rules are
guaranteed to be resolvable, given that the construction of canonical
paths enforces coincidence with the endpoints stated in the type.

It is highly likely that we shall also want an $\eta$-law.
\[
\textsc{typ-$\eta$}\;\;
  R=S\tp\star T\;\Rightarrow\; R\ni r \jq \lambda i.\,(r\hb R)\:i
\]

However, the key thing we can do with types joined by paths is to
transport values between them, along those paths. Transportation is
not teleportation. Rather, it can be seen as the extrusion of a value
from one point on a type path to every point on the type path,
creating a path between values, followed by the projection of the
value at a chosen destination. Before we consider transportation, let
us introduce paths between values.


\section{Paths Between Values Over Type Paths}

If we have a path connecting types, we establish one basis for
proposing connections between values of those types.
\newcommand{\vp}[1]{-\!\!\!\left[#1\right]\!\!\!-}
\[\textsc{chk-vap}\;
\Rule{\star'\ni S\tp\star T \quad S\tp\star T\ni P\quad
      S\ni s\quad T\ni t}
     {\star'\ni s \vp{P\hb S\tp\star T} t}
\]
To construct such a path, one may exhibit the points thereon.
\[\textsc{chk-vap-lam}\;
\Rule{\begin{array}{c}
    \bar i\vdash  (P\hb S\tp\star T)\:i\ni p[i]\\\
      S\ni s\jq p[\pl]\quad T\ni p[\pr]\jq t \end{array}}
     {s \vp{P\hb S\tp\star T} t \ni \lambda i.\,p[i]}
\]
Of course, we must also have projection.
\[\begin{array}{c}
\textsc{syn-vap-app}\;
\Rule{e\in s \vp{P\hb S\tp\star T} t\quad \pt a}
     {e\:a\in(P\hb S\tp\star T)\:a}
\\
\textsc{red-vap-$\beta$}\;
(\lambda i.\,p[i]:s \vp{P\hb S\tp\star T} t)\:a\leadsto p[a]:(P\hb S\tp\star T)\:a\\
\textsc{red-vap-$\pl$}\;
  e\in s \vp{P\hb S\tp\star T} t \;\Rightarrow\; e\:\pl\leadsto s\hb S\\
\textsc{red-vap-$\pr$}\;
  e\in s \vp{P\hb S\tp\star T} t \;\Rightarrow\; e\:\pr\leadsto t\hb T\\
\textsc{vap-$\eta$}\;\;
  R=s \vp{P\hb S\tp\star T} t\;\Rightarrow\; R\ni r \jq \lambda i.\,(r\hb R)\:i
\end{array}\]


\section{Rocketing Between Types}

We can transport values along paths, and not just from
one end to the other, but between \emph{any} connected points,
allowing us to ensure that values transported along type paths
remain connected by value paths. In OTT, the `coercion' operation
really did teleport values, and its `coherence' (the equality of
teleported values before and afterwards) was asserted by
brute force, exploiting proof irrelevance. Here, values travel
in a rocket, passing continuously through space along a given type
path.
\newcommand{\xp}[4]%
  {#1\:\overline{\underline{\left>#2\!\!\mid\!\! #3 \!\!\mid\!\! #4\right>\!\!}}\;}
\[
\Rule{\pt a\quad  E\in S\tp\star T \quad \pt b\quad E\:a\ni r}
     {\xp r a E b \in E\:b}
\]
The notation is intended to suggest sending $r$ from $a$ to $b$ along
$E$, but as far as typing is concerned, this construct is an
elimination form for $E$, rather than $r$. We shall ensure that
\[
s\vp{E\hb S\tp\star T} \xp s \pl E \pr \ni
\lambda i.\,\xp s \pl E i
\]
The typical point on the path certainly has the right type, and it is
surely connected at the $\pr$ end, but to check the connection at the
$\pl$ end, we shall need that
\[
  S\ni s\jq \xp s \pl E \pl
\]

Rocketing must compute, so for every means we have to introduce
paths between types, we must be able to explain the behaviour of a
rocket along those paths. Moreover, for every equation which
holds judgmentally between type paths, we must ensure that the
rockets they induce coincide also.

However, we
shall have, in general, that remaining motionless changes nothing.
\[
\textsc{go-stop}\;\;\xp s a E a \leadsto s\hb E\:a
\]
That is, we cannot change anything by standing still on a
path. Moreover, we cannot change anything by moving along a path
if the path itself is degenerate.
\[
\textsc{go-still}\;\;\xp s a {\lambda\_.\,S} b \leadsto s\hb S
\]
Note that in the presence of $\eta$, a `pattern' $\lambda i.\,P[i]$
matches some $T$ if it the body of the $\eta$-expansion of $T$ over
some fresh variable $i$ matches $P[i]$. The above rule effectively
demands an occur-check.

There may very well be circular paths which are not degenerate, but in
some sense, they must move away from their origin and then return,
rather than remaining constantly in one place. Moreover, to achieve
any noticeable difference by transmitting a value along such a
circular path, one must travel a nontrivial distance. The above rules
thus do not, conceptually at least, compromise univalence, nor do they
presume it.

\section{Rocketing for Structural Type Paths}

We say that a type path $\lambda i.\,P[i]$ is `structural' if $P[i]$
is given by a canonical type former, meaning that any movement in the
path happens within the components of that type.  The idea behind OTT
is that the only closed paths are the structural paths, but univalence
adds others. Whatever other paths we may have, we shall certainly have
the structural paths, so at least these must admit rocketing. For base
types, $\Zero$, $\One$, $\Two$, structural paths are degenerate and
their rockets act as the identity.

For dependent pairs, a structural path must be componentwise. Note
that there are paths between pair types which are not componentwise
(e.g., swapping the components of a binary product), but that these
paths cannot be given structurally. A rocket of pairs is a pair of
rockets. Eliding obvious type information, we get
\[\begin{array}{l}
\xp {(s,t)} a{\lambda i.\,(x\hb S[i])\times T[i,x]}b \leadsto
  (\hat s[b], \hat t[b]):\cdots\;\;\textrm{where} \\
\qquad\begin{array}{l}
  \hat s[j]= \xp s a {\lambda i.\,S[i]} j \\
  \hat t[j]= \xp t a {\lambda i.\,T[i,\hat s[i]]} j \\
\end{array}
\end{array}\]
If we have $\eta$ for pairs, then pattern matching a pair
means pattern matching its expansion, i.e., projecting components,
rather than waiting for the constructor.
Observe that we construct $\hat s$ and $\hat t$, being the images of $s$,
and $t$ over the entire path. Moreover, we need the full generality of
$\hat s$ to give the path over which $t$ can rocket. Note that $\hat
s[s]$ is just $s$ by the \textsc{go-stop} rule, so $t$ has the
correct type for its rocket. We must also check critical pairs: if
$b$ is $a$, then we do indeed get $(s,t)$ back; if the whole path is
degenerate, then so is its $S[\_]$ component, which makes $\hat s$ a
constant $s$, and the second component path also degenerate $T[\_,s]$.

For dependent functions, we rocket back and forth.
\[\begin{array}{l}
\xp{(\lambda x.\:t[x])}a{\lambda i.\,(x\hb S[i])\to T[i,x]}b \leadsto
  \lambda x.\, \hat t[b]:\cdots\;\;\textrm{where} \\
\qquad\begin{array}{l}
  \hat s[j]= \xp x b {\lambda i.\,S[i]} j \\
  \hat t[j]= \xp {(t[\hat s[a]]} a {\lambda i.\,T[i,\hat s[i]]} j \\
\end{array}
\end{array}\]
As before, if we have $\eta$ for functions, then the $\lambda$-pattern
forces expansion. We can readily check that $r$ (which comes from the
$a$ side) gets an argument it can use, and that the eventual output (on the $b$ side),
has the required type $T[b,\hat s[b]] = T[b,x]$, again by the
\textsc{go-stop}
rule. Again, if $a$ and $b$ coincide or the path is degenerate, this
rocket amounts to the identity.


%\appendix
%\section{Appendix Title}
%
%This is the text of the appendix, if you need one.

\acks

Many acknowledgments are due. We have had a lot of assistance, much
inspiration from colleagues and many enlightening conversations. Names
will be named and honour bestowed if and when this paper ceases to be
anonymous.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\bibliography{Crossroads}

%\begin{thebibliography}{}
%\softraggedright
%
%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
%P. Q. Smith, and X. Y. Jones. ...reference text...
%
%\end{thebibliography}


\end{document}

%                       Revision History
%                       -------- -------
%  Date         Person  Ver.    Change
%  ----         ------  ----    ------

%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

